Measurement Clusterization in D-optimal Designs for Bayesian Linear Inverse Problems

### 2023-05-05 latex/vanishing.tex
Moved example to appendix, added questions and answers in intro.

\section{D-Optimal Designs when $\modcov = 0$}\label{section:vanishing}
We will now prove Theorem \ref{thm:char} which caracterizes D-optimal
designs when $\modcov = 0$. The necessary first-order condition for
D-optimality of Theorem \ref{thm:constrained} for $\modcov = 0$
become:

@@

\begin{proposition}\label{prop:true target}
  Let $\obs$ with $m$ observations satisfy the nonlinear eigenvalue
  problem \eqref{eq:eigenproblem}. Let $\{\eta_i\}_{i=1}^{\infty}$
  eigenvalues of $\obs^*\obs$ and $\{\lambda_i\}_{i=1}^{\infty}$ the
  corresponding eigenvalues of $\fwd \prcov \fwd^*$. Let $k:=\rank
  \obs^*\obs$. Without loss of generality, let $\eta_i > 0$ for $i\leq
  k$ and $\eta_i = 0$ for $i > k$. Then:
  \begin{enumerate}
    \item $k \leq m$ and $\obs^*\obs$ has exactly $k$ positive
      eigenvalues.
    \item
      \begin{equation*}
        \tar(\obs) = \frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i) = \frac12 \sum_{i=1}^{m} \log (1 + \sigma^{-2}\lambda_i\eta_i).
      \end{equation*}
    \item Furthermore, if $\obs$ isD-optimal, $\eta_i > 0$ for
      eigenvectors corresponding to the $k$ largest $\lambda_i$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  Part (1) is trivial. To see part (2) holds: 
  \begin{align}
    \begin{split}
      \tar(\obs) &= \frac12\log \det \left (I + \sigma^{-2} \prcov^{1/2} \fwd ^* \obs^*
      \obs \fwd \prcov^{1/2}\right )\\% \text{ (by definition)}\\
      %
      &= \frac12 \log \det \left (I + \sigma^{-2} \obs^* \obs \fwd
      \prcov\fwd^* \right ) \text{ (Sylvester's Determinant
      Theorem)}\\
      %
      %
      %
      &=\frac12 \log \prod_{i=1}^{\infty} ( 1 + \sigma^{-2} \lambda_i\eta_i ) \text{ (Proposition \ref{prop:same ev})} \\
      %
      %
      %
      %&=\frac12 \log \left ( \prod_{i=1}^{k} ( \lambda_i^{-1} + \sigma^{-2} \eta_i )\prod_{i=1}^{k} \lambda_i \right )
      %
      %
      %
      &=\frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i). 
      %
      %
      %% &=\frac12 \sum_{i=1}^{k} \log(\sigma^2\lambda_i^{-1} + \eta_i) - k\log \sigma + \frac12\sum_{i=1}^k \log \lambda_i.
    \end{split}
  \end{align}
  Part (3) holds since $\log$ is increasing and $\eta_i \geq 0$.
\end{proof}

@@

\begin{proof}
  Part (1) is immediate for any observation operator $\obs$ that
  satisfies the unit norm constraint on observation vectors. Part (2)
  was proved in Proposition \ref{prop:same ev}. Part (3) was proved in
  Proposition \ref{prop:true target}.

@@

    \end{axis}
  \end{tikzpicture}
  \caption{A comparison of the eigenvalues of the pushforward
    posterior precision $(\fwd\prcov\fwd^*)^{-1} +
    \sigma^{-2}\obs^*\obs$ for a D-optimal design (left) and a
    sub-optimal design (right). Both designs are allowed $m=3$
    observations. We assume $\sigma^2=1$ and thus, the blue area has
    accumulated height of $\sigma^{-2}m = 3$ in both plots. The
    D-optimal design (left) increases precision where it is
    lowest. The sub-optimal design (right) does not.}
  \label{fig:optimal vs not}
\end{figure}

Theorem \ref{thm:char} facilitates understanding of D-optimal designs
when $\modcov = 0$: Imagine each eigenvector of $\fwd \prcov \fwd^*$
corresponds to a graduated lab cylinder (Figure
\ref{fig:cylinder}). Each cylinder $i$ is filled, a-priori, with green
liquid of $\lambda_i^{-1}$ volume units. We are allowed $m$
observations, so we have blue liquid of volume $\sigma^{-2}m$ units at
our disposal. When we seek a D-optimal design, we distribute the blue
liquid by repeatedly adding a drop to whatever cylinder currently has
the lowest level of liquid in it, as long as its index $i \leq m$. The
result of such a procedure is illustrated in Figure \ref{fig:optimal
  vs not}.

\begin{figure}%{r}{0.25\textwidth} 
    %\centering
    \includegraphics[width=5cm, height=5cm]{cylinders.jpg}
    \caption{Three graduated lab cylinders, corresponding to three
      eigenvectors. Prior eigenvalues not shown.}
    \label{fig:cylinder}
\end{figure}

\section{Clusterization poses no obstruction to D-optimality}\label{section:clusterization}
The necessary conditions for D-optimality are (1), (2) and (4) of
Theorem \ref{thm:char}. These conditions, however, still leave some
freedom in choosing $\obs$. This freedom may give rise to
clusterization, as illustrated in Figure \ref{fig:clusterization} and
demonstrated below.

Consider an inverse problem for which $\lambda_1^{-1} = 0.1,
\lambda_2^{-1} = 0.2,\sigma^2=1$ and $\{\ev_i\}_{i=1}^\infty$ are
eigenvectors of $\fwd\prcov\fwd^*$. Two D-optimal designs $\obs_1,
\obs_2$ are presented in \eqref{eq:two designs}, and it is easy to
verify that $\obs_1^*\obs_1 = \obs_2^*\obs_2$. However, $\obs_1$
exhibits clusterization ($\meas_3 = \meas_5 = \ev_1$), while $\obs_2$
does not. Both of $\obs_1$ and $\obs_2$ are D-optimal, and we conclude
that clusterization poses no obstruction to D-optimality.

\begin{equation}\label{eq:two designs}
  \obs_1 =
  \begin{pmatrix}
    \sqrt{\frac{11}{40}}\ev_1 - \sqrt{\frac{29}{40}}\ev_2 \\
    \sqrt{\frac{11}{40}}\ev_1 + \sqrt{\frac{29}{40}}\ev_2 \\
    \ev_1 \\
    \ev_2 \\
    \ev_1 \\    
  \end{pmatrix},
  \obs_2 =
  \begin{pmatrix}
    \sqrt{\frac{3}{8}}\ev_1 - \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{3}{8}}\ev_1 + \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 - \sqrt{\frac{3}{5}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 + \sqrt{\frac{3}{5}}\ev_2 \\
    \ev_1 \\    
  \end{pmatrix}
\end{equation}

A computer implementation of Lemma \ref{thm:char} (code in
Supplementary material) generates $\obs_1$ as a solution to the
D-optimal design problem mentioned in the previous
paragraph. Randomized numerical simulations (see code in Supplementary
material) give rise to D-optimal designs that exhibit clustrization
more than 95\% of the time. We conclude that clusterization poses no
obstruction to D-optimality for the model presented in this study and
that sensor clusterization is generic.

>>>>

\section{Characterizing D-Optimal Designs when $\modcov = 0$}\label{section:vanishing}
We will now prove Theorem \ref{thm:char} which caracterizes D-optimal
designs when $\modcov = 0$. The necessary first-order condition for
D-optimality of Theorem \ref{thm:constrained} for $\modcov = 0$
become:

@@

\begin{proposition}\label{prop:true target}
  Let $\obs$ with $m$ measurements satisfy the nonlinear eigenvalue
  problem \eqref{eq:eigenproblem}. Let $\{\eta_i\}_{i=1}^{\infty}$
  eigenvalues of $\obs^*\obs$ and $\{\lambda_i\}_{i=1}^{\infty}$ the
  corresponding eigenvalues of $\fwd \prcov \fwd^*$. Let $k:=\rank
  \obs^*\obs$. Without loss of generality, let $\eta_i > 0$ for $i\leq
  k$ and $\eta_i = 0$ for $i > k$. Then:
  \begin{enumerate}
    \item $k \leq m$ and $\obs^*\obs$ has exactly $k$ positive
      eigenvalues.
    \item
      \begin{equation*}
        \tar(\obs) = \frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i) = \frac12 \sum_{i=1}^{m} \log (1 + \sigma^{-2}\lambda_i\eta_i).
      \end{equation*}
    \item Furthermore, if $\obs$ isD-optimal, $\eta_i > 0$ for
      eigenvectors corresponding to the $k$ largest $\lambda_i$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  Part (1) is trivial. To see part (2) holds: 
  \begin{align}
    \begin{split}
      \tar(\obs) &= \frac12\log \det \left (I + \sigma^{-2} \prcov^{1/2} \fwd ^* \obs^*
      \obs \fwd \prcov^{1/2}\right )\\% \text{ (by definition)}\\
      %
      &= \frac12 \log \det \left (I + \sigma^{-2} \obs^* \obs \fwd
      \prcov\fwd^* \right ) \text{ (Sylvester's Determinant
      Theorem)}\\
      %
      %
      %
      &=\frac12 \log \prod_{i=1}^{\infty} ( 1 + \sigma^{-2} \lambda_i\eta_i ) \text{ (Proposition \ref{prop:same ev})} \\
      %
      %
      %
      %&=\frac12 \log \left ( \prod_{i=1}^{k} ( \lambda_i^{-1} + \sigma^{-2} \eta_i )\prod_{i=1}^{k} \lambda_i \right )
      %
      %
      %
      &=\frac12 \sum_{i=1}^{k} \log (1 + \sigma^{-2}\lambda_i\eta_i). 
      %
      %
      %% &=\frac12 \sum_{i=1}^{k} \log(\sigma^2\lambda_i^{-1} + \eta_i) - k\log \sigma + \frac12\sum_{i=1}^k \log \lambda_i.
    \end{split}
  \end{align}
  Part (3) holds since $\log$ is increasing and $\eta_i \geq 0$.
\end{proof}

@@

\begin{proof}
  Part (1) is immediate for any measurement operator $\obs$ that
  satisfies the unit norm constraint on measurement vectors. Part (2)
  was proved in Proposition \ref{prop:same ev}. Part (3) was proved in
  Proposition \ref{prop:true target}.

@@

    \end{axis}
  \end{tikzpicture}
  \caption{A comparison of the eigenvalues of the pushforward
    posterior precision $(\fwd\prcov\fwd^*)^{-1} +
    \sigma^{-2}\obs^*\obs$ for a D-optimal design (left) and a
    sub-optimal design (right). Both designs are allowed $m=3$
    measurements. We assume $\sigma^2=1$ and thus, the blue area has
    accumulated height of $\sigma^{-2}m = 3$ in both plots. The
    D-optimal design (left) increases precision where it is
    lowest. The sub-optimal design (right) does not.}
  \label{fig:optimal vs not}
\end{figure}

Theorem \ref{thm:char} facilitates understanding of D-optimal designs
when $\modcov = 0$: Imagine each eigenvector of $\fwd \prcov \fwd^*$
corresponds to a graduated lab cylinder (Figure
\ref{fig:cylinder}). Each cylinder $i$ is filled, a-priori, with green
liquid of $\lambda_i^{-1}$ volume units. We are allowed $m$
measurement, so we have blue liquid of volume $\sigma^{-2}m$ units at
our disposal. When we seek a D-optimal design, we distribute the blue
liquid by repeatedly adding a drop to whatever cylinder currently has
the lowest level of liquid in it, as long as its index $i \leq m$. The
result of such a procedure is illustrated in Figure \ref{fig:optimal
  vs not}.

\begin{figure}%{r}{0.25\textwidth} 
    \centering
    \includegraphics[width=5cm, height=5cm]{cylinders.jpg}
    \caption{Three graduated lab cylinders, corresponding to three
      eigenvectors. Prior eigenvalues not shown.}
    \label{fig:cylinder}
\end{figure}

\section{Clusterization poses no obstruction to D-optimality}\label{section:clusterization}T
The necessary conditions for D-optimality of Theorem \ref{thm:char}
leave some freedom in choosing $\obs$, which may or may not give rise
clustered designs. Consider an inverse problem for which the first two
eigenvalues of $\fwd \prcov\fwd^*$ are $\lambda_1^{-1} = 0.1,
\lambda_2^{-1} = 0.2$, its eigenvectors are $\{\ev_i\}_{i=1}^\infty$
and $\sigma^2=1$. Consider the following $\obs_1, \obs_2$ (see also
Figure \ref{fig:clusterization}):
\begin{equation}\label{eq:two designs}
  \obs_1 =
  \begin{pmatrix}
    \sqrt{\frac{11}{40}}\ev_1 - \sqrt{\frac{29}{40}}\ev_2 \\
    \sqrt{\frac{11}{40}}\ev_1 + \sqrt{\frac{29}{40}}\ev_2 \\
    \ev_1 \\
    \ev_2 \\
    \ev_1 \\    
  \end{pmatrix},
  \obs_2 =
  \begin{pmatrix}
    \sqrt{\frac{3}{8}}\ev_1 - \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{3}{8}}\ev_1 + \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 - \sqrt{\frac{3}{5}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 + \sqrt{\frac{3}{5}}\ev_2 \\
    \ev_1 \\    
  \end{pmatrix}
\end{equation}
\noindent It is easy to verify that $\obs_1^*\obs_1 = \obs_2^*\obs_2$
and that both $\obs_1$ and $\obs_2$ are D-optimal. However, $\obs_1$
is a clustered design ($\meas_3 = \meas_5 = \ev_1$), while $\obs_2$ is
not.

We are now ready to answer some of the questions posed in the
introduction:

\textbf{Question \ref{q:why}}: Since $\obs_1$ is clustered and
D-optimal, measurement clusterization poses no obstruction to
D-optimality. Furthermore, a computer implementation of Lemma
\ref{thm:char} also generates $\obs_1$ as a solution to the D-optimal
design problem. Randomized numerical simulations give rise to
D-optimal designs that exhibit clustrization more than 95\% of the
time (see code in supplementary). Since our model is as generic as
possible, we conclude that it is possible that measurement
clusterization is a generic phenomenon that.

\textbf{Question \ref{q:replace}}: Since $\obs_2$ is D-optimal but not
clustered, we conclude that it is possible to substitute a clustered
design for an equally optimal non-clustered design. Achieveing this in
the setup of our model not hard, but in real life problems this may
prove more difficult.
